### 2017-01-26 manuscript.rmd
small tweaks to text

Although there are mature methods and tools for statistical analysis of character evolution [@Garamszegi:2014ew] and gene expression [@Robinson:2009cw], Levin *et al.* [@Levin:2016bp] drew on neither of these active areas of research and instead applied an *ad hoc* pairwise comparison of ortholog gene expression between species. They characterized each gene in each species as having expression that peaks in early, mid, or late temporal phase of development. The goodness of fit to these patterns was not considered, and alternative patterns, such as the possibility that there are more phases of expression or no distinct phases of expression, were not evaluated. For each species pair, they identified the orthologs shared by these species. This list of genes was different from species pair to species pair). They then calculated a similarity score for each temporal phase for each species pair based on the fraction of genes that exhibited the same patterns in each species. The distributions of similarity scores are plotted in their [Figure 4d](http://www.nature.com/nature/journal/v531/n7596/fig_tab/nature16994_F4.html), and their Kolmogorov–Smirnov (KS) tests indicated that the early distribution and late distribution were each significantly different from mid distribution (P < 10<sup>-6</sup> and P < 10<sup>-12</sup>, respectively). This is the support they presented for the inverse hourglass model.

@@

We previously described some concerns with their interpretations of these results [@Hejnol:2016hw]. The main issue is that because they only sampled one species per phylum, there is no way to tell if observed differences are unique to the branches that gave rise to clades designated as phyla or occurred along the many other branches in the tree. Here we examine the support for the inverse hourglass model itself. We focus on the matrix of pairwise comparisons that their KS tests and their Figure 4d are based on. As this figure shows (regenerated here as our Figure XXLevin a), there is not a strong distinction between these temporal phases - the distributions for the early, mid, and late distributions overlap considerably. The medians are close in value, and the range of mid phase includes the entire range of the early phase. Further inspection (Figure XXLevin a) reveals that the mid phase distribution has several outliers with very low similarity scores.

The first problems we identified were general stastistical issues that are not problems with pairwise comparisons *per se*. First, in the published analyses every data point was included twice because both reciprocal comparisons (which have the same values) were retained. For example, there is both a nematode to arthropod comparison and an arthropod to nematode comparison. As a consequence, there are 90 entries for the 45 pairwise comparisons, and by doubling the data the significance of the result appears stronger than it actually is. After removing the duplicate values, the p values are far less significant, 0.002 for the early-mid comparison and on the order of 10<sup>-6</sup> for early-late. Second, the test they used (KS test) is not appropriate for the hypothesis they seek to evaluate. The KS test does not just evaluate whether one distribution is greater than the other, it also tests whether the shape of the distributions are the same. In addition, the samples in this dataset are matched (*i.e.*, for each pairwise comparison there is a early, mid, and late expression value), which the KS test does not take into account. The Wilcoxon test is instead appropriate in this case.

The next issue we found is a general problem with pairwise comparisons. Inspecting all five of the lowest values in the mid phase distribution (Figure XXLevin a) are for pairwise comparisons that include the ctenophore (comb jelly). When the nine pairwise comparisons that include the ctenophore are removed and the general statistical issues noted above are addressed, there is no significant difference between the early phase and mid phase distributions (P = 0.1428 for the early-mid comparison and P < 10<sup>-5</sup> for the late-mid comparison). The inverse hourglass turns into a bottle (Figure XXLevin b).

This highlights a well understood property of pairwise comparisons across species [@Felsenstein:1985ua]: evolutionary changes along a given branch, like those along the ctenophore branch, impact each of the multiple pairwise comparisons that includes that branch. The pairwise comparisons are therefore not independent - one evolutionary change impacts multiple comparisons. This can give the impression of a general pattern across the tree that is instead specific to changes along one part of the tree. The number of comparisons impacted by each change depends on the structure of the phylogenetic tree, *i.e.* how the species are related to each other.

Phylogenetic comparative methods were developed specifically to address this problem [@Felsenstein:1985ua]. While we do demonstrate it impacts the Levin *et al.* analysis, we did not perform a phylogenetic reanalysis of this study, as we did for the KMRR study. This is because the similarity metric computed in the pairwise comparisons of Levin *et al.* are based on different genes for different species pairs. This means that they do not generate a trait that can be evaluated across a phylogeny, as KMRR's Tau can be. A full phylogenetic reanalysis would be possible by using upstream analysis products to re-derive new expression summary statistics suitable for phylogenetic comparative methods.

>>>>

Levin *et al.* [@Levin:2016bp] arrived at this conclusion my making pairwise comparisons of ortholog expression through development between species. They characterized each gene in each species as having expression that peaks in early, mid, or late temporal phase of development. The goodness of fit to these patterns was not considered, and alternative patterns, such as the possibility that there are more phases of expression, were not evaluated. For each species pair, they identified the orthologs shared by these species. This list of shared genes was different from species pair to species pair. They then calculated a similarity score for each temporal phase for each species pair based on the fraction of genes that exhibited the same patterns in each species. The distributions of similarity scores are plotted in their [Figure 4d](http://www.nature.com/nature/journal/v531/n7596/fig_tab/nature16994_F4.html), and their Kolmogorov–Smirnov (KS) tests indicated that the early distribution and late distribution were each significantly different from mid distribution (P < 10<sup>-6</sup> and P < 10<sup>-12</sup>, respectively). This is the support they presented for the inverse hourglass model.

@@

We previously described some concerns with their interpretations of these results [@Hejnol:2016hw]. The main concern was that because they only sampled one species per phylum, there is no way to tell if observed differences are unique to the branches that gave rise to clades designated as phyla or occurred along the many other branches in the tree. Here we examine the support for the inverse hourglass model itself. We focus on the matrix of pairwise comparisons that their KS tests and their Figure 4d are based on.

The first problems we identified were general stastistical issues that are not problems with pairwise comparisons *per se*. First, in the published analyses every data point was included twice because both reciprocal comparisons (which have the same values) were retained. For example, there is both a nematode to arthropod comparison and an arthropod to nematode comparison. As a consequence, there are 90 entries for the 45 pairwise comparisons, and by doubling the data the significance of the result appears stronger than it actually is. After removing the duplicate values, the p values are far less significant, 0.002 for the early-mid comparison and on the order of 10<sup>-6</sup> for early-late. Second, the test they used (KS test) is not appropriate for the hypothesis they seek to evaluate. The KS test does not just evaluate whether one distribution is greater than the other, it also tests whether the shape of the distributions are the same. In addition, the samples in this dataset are matched (*i.e.*, for each pairwise comparison there is a early, mid, and late expression value), which the KS test does not take into account. The Wilcoxon test is instead appropriate in this case. When applied to the de-duplicated data, the significance of this test is 0.02 for the early-mid comparison and on the order of 10<sup>-7</sup> for early-late.

The next issue we found is a general problem with pairwise comparisons. We found that all five of the lowest values in the mid phase distribution (Figure XXLevin a) are for pairwise comparisons that include the ctenophore (comb jelly). When the nine pairwise comparisons that include the ctenophore are removed and the general statistical issues noted above are addressed, there is no significant difference between the early phase and mid phase distributions (P = 0.1428 for the early-mid comparison and P < 10<sup>-5</sup> for the late-mid comparison). The inverse hourglass turns into a bottle (Figure XXLevin b).

This highlights a well understood property of pairwise comparisons across species [@Felsenstein:1985ua]: evolutionary changes along a given branch, like those along the ctenophore branch, impact each of the multiple pairwise comparisons that includes that branch. The pairwise comparisons are therefore not independent - different pairwise comparisons are impacted by canges along some of the same branches. This can give the impression of a general pattern across the tree that is instead specific to changes along one part of the tree. The number of comparisons impacted by each change depends on the structure of the phylogenetic tree, *i.e.* how the species are related to each other.

Phylogenetic comparative methods were developed specifically to address this problem [@Felsenstein:1985ua]. While we do demonstrate this problem impacts the Levin *et al.* analysis, we did not perform a phylogenetic reanalysis of this study, as we did for the KMRR study. This is because the similarity metric computed in the pairwise comparisons of Levin *et al.* are based on different genes for different species pairs. This means that they do not generate a trait that can be evaluated across a phylogeny, as KMRR's Tau can be. A full phylogenetic reanalysis would be possible by using upstream analysis products to re-derive new expression summary statistics suitable for phylogenetic comparative methods.
